A new tidy data structure to support exploration and modeling of temporal data

### 2019-08-11 tsibble.Rmd
done with revision for subsection on index

Figure \@ref(fig:tsibble-semantics) sketches out the data form required for a tsibble, the same format as the "tidy data". Beyond the layout, tsibble gives the contextual meaning to variables in order to construct the temporal data object, as newly introduced "index" and "key" semantics stated in 1 and 2 above. As a result of the definitions 3 and 4, tsibble is *tidier* than "tidy data", placing itself as a data object that is also suitable for modeling.

@@

To demonstrate the concept of the tsibble, Table \@ref(tab:tb-sub) presents a subset of tuberculosis cases estimated by @tb-data. It contains 12 observations and 5 variables arranged in a "longer" tabular form. Each observation comprises the number of people who are diagnosed with tuberculosis for each gender at three selected countries in the years of 2011 and 2012. To turn this data into a tsibble: (1) column `year` is declared as the index variable; (2) the key is specified to consist of columns `country` and `gender`. The column `count` is the only measured variable in this data, but the structure is sufficiently flexible to hold other measured variables; for example, adding the corresponding population size (if known) in order to normalize the count later.

@@

(ref:tb-sub) A small subset of estimates of tuberculosis burden generated by World Health Organization in 2011 and 2012, with 12 observations and 5 variables. The index refers to column `year`, the key to multiple columns: `country` and `gender`, and the measured variable to column `count`.

The new data structure, tsibble, bridges the gap between raw temporal data and model inputs. Contextual semantics are introduced to tidy data in order to support more intuitive time-related manipulations and enlighten new perspectives for time series model inputs. Index, key and time interval are the three stone pillars to this new semantically structured temporal data. Each is now described in more detail.

@@

Time provides a contextual basis for temporal data. A variable representing time is essential for a tsibble, and is referred to as an "index". The "index" is an intact data column rather than a masked attribute, which makes time visible and accessible to users. This is highly advantageous when manipulating time. For example, one could easily extract time components, such as time of day and day of week, from the index to visualize seasonal effects of response variables. One could also join other data sources to the tsibble based on common time indexes. The accessibility of the tsibble index motivates data analysis towards transparency and human readability. When the "index" is available only as meta information (such as in the `ts` class), it creates an obstacle for analysts to write these simple queries in a programmatic manner, which should be discouraged from an analytical point of view.

Many different representations of time can be spotted in the wild. A date-time object, universally accepted across computing systems, is the most commonly used type for representing time. Date-time also typically associates with a time zone including adjustments such as daylight saving time. This diversity and time zone is acknowledged and accommodated by tsibble's index. When creating a tsibble, time indices are arranged from past to future within each series for the strict temporal ordering that is assumed by time series operations.

>>>>

Figure \@ref(fig:tsibble-semantics) sketches out the data form required for a tsibble, an equivalent of the "tidy" format. Beyond the layout, tsibble gives the contextual meaning to variables in order to construct the temporal data object, as newly introduced "index" and "key" semantics stated in the first and second definitions above. Variables other than "index" and "key" are considered as measurements. The third and fourth definitions imply that tsibble is *tidier* than "tidy data", placing itself as a model input that gives rise to more robust and reliable downstream analytics.

@@

To materialize the abstraction of the tsibble, a subset of tuberculosis cases estimated by @tb-data presented in Table \@ref(tab:tb-sub) is used as an example. It contains 12 observations and 5 variables landing in a tidy data form. Each observation comprises the number of people who are diagnosed with tuberculosis for each gender at three selected countries in the years of 2011 and 2012. From tidy data to tsibble data, "index" and "key" should be declared: column `year` as the "index" variable, and column `country` together with `gender` as the "key" variable forming the observational units. Column `count` is the only measured variable in this data, but the data structure is sufficiently flexible to hold more measurements; for example, slotting the corresponding population size (if known) into the data column for normalizing the count later. Note this data further satisfies the must-have for the distinct rows determined by index and key, and regularly spans over one-year interval.

@@

(ref:tb-sub) A small subset of estimates of tuberculosis burden collected by World Health Organization in 2011 and 2012, with 12 observations and 5 variables. The index refers to column `year`, the key to multiple columns: `country` and `gender`, and the measured variable to column `count`.

The new tsibble structure bridges the gap between wild temporal data and the tamed state of model inputs. The proposed contextual semantics is the new addition to tidy data in order to support more intuitive time-related manipulations and enlighten new perspectives for time series and panel model inputs. Index, key and time interval build the three stone pillars to this new semantically structured temporal data. Each is now described in more detail.

@@

> *Index is a variable with inherent ordering from past to present.*

Time provides the contextual basis for temporal data. Time can be seen in numerous representations, from sequential numerics to the most commonly accepted date-times. Regardless of this diversity, time should be inherently ordered from past to present, so should the index variable to a tsibble.

Index is an explicit data variable rather than a masked attribute (such as in the `ts` and `zoo` classes), exposing a need for more accessible and transparent time operations. It is often to visualize and model seasonal effects of measurements of interest, meaning that time components such as time of day and day of week from the index should be easily extracted upon request. But when the index is available only as meta information, an obstacle is inevitably created for analysts to write such simple queries in an ad hoc and programmatic manner. From an analytical point of view it should be discouraged.
